To be FAIR: Theory Specification Needs an Update

### 2025-03-05 manuscript.Rmd
consistent discrimination of vague and ambiguous (striking empierical content)

The so-called "replication crisis" has prompted extensive scientific reforms [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that the abundance of non-replicable findings 
was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses by commiting to an analysis plan in advance.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
The difference between these definitions becomes clear when considering the act of changing an infant's dirty diaper.
Under the original definition, one might hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
Under the expanded definition, one might hypothesize the opposite, as the act may fulfill the need for relatedness.

Scholars have raised concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
One main concern is that theories lack *formalization* [@szollosiArrestedTheoryDevelopment2021].
<!-- I still believe formalization != precision. Theories can be 100% formal and still 0% precise. -->
<!-- NvD: We should add somewhere what we mean by "formalization." In philosophy of science, formal logic about syntax (the propositions and predicates can have any meaning). However, what we mean, I think, is the explication of a narrative in (something so detailed and explicit that it can be captured with little error in) analytical or computational form (but the meaning remains. E, M, and C stand form something is E = MC<sup>2</sup>).   -->
When theories are difficult to understand without either subjective interpretation or additional background knowledge,
it becomes difficult do derive precise predictions,
and therefore hard to falsify the theory. 
A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
Despite these concerns, scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and improvement have been largely overlooked^[Here, we use "induction" to refer to inferences from specific observations to general theories. Others have used the term "abduction" to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960]. For present purposes, the terms are interchangeable.].
<!-- CvL considers abduction to be a special case of induction, because defining what makes a theory "best" or "plausible" requires making auxiliary assumptions, which restrict the space of potential inductive inferences. Peirce also stated that abduction "is not essentially different from induction" [@peirceCollectedPapersCharles1960, p.4881].  NvD dissents: -->
<!-- NvD: I cannot accept this footnote as an adequate characterization. Abduction is not just "inference to the best explanation." It is typically used as inference to a plausible explanation (but there could be more). See for instance Brian Haig's 2005 paper. Also, I don't see why it matters either way "why defining what makes a theory "best" requires auxiliary assumptions". At the very least I would like to change "we consider abduction to be a special case of induction", because I disagree. I am fine with the last sentence though ;-) My suggestion is to write the following: -->
<!-- NvD: We use induction to describe inferences from specific observations to general theories. In philosophy of science the term "abduction" is often used for the inference from observations to plausible explanations. We note that induction and abduction come in several varieties in philosophy science, which overlap in several respects For present purposes, however, the terms are interchangeable. -->
<!-- Peirce: CP 8.233 Cross-Ref:†† -->
<!-- 233. This is not essentially different from induction. Only it is most -->
<!-- usually an induction from instances which are not discrete and numerable. I now -->
<!-- call it Qualitative Induction. It is this which I used to confound with the second -->
<!-- line of procedure, or at least not to distinguish it sharply. -->
The present paper addresses this knowledge gap by applying, for the first time,
open science principles to psychological theory.
We apply introduce *FAIR theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use. 
<!-- NvD: Maybe we should add "specification" as part of "formalization" or "formalisation" as a part of "specification", because not everybody that is advocating better theories thinks that we should focus on formalization.  -->
More rigorous formalization increases theories' *empirical content* [@popperLogicScientificDiscovery2002] because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Compared to theories expressed in natural language,
formal theories facilitate inconsistency checking and evaluation of a theory's (lack of) ambiguity.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.

@@

Making theory FAIR allows researchers to more easily find a relevant framework;
access and understand it; interact with it in a practical sense, for example, by deriving predictions from it, or using it to select control variables; and reuse it, contributing changes to existing theories or splitting of in a new direction.
Whereas the idea of theory can be quite nebulous to empirical psychologists,
FAIR theory makes theoretical work practical and tangible, incorporating theory into scholars' workflows.
Having a concrete object to iterate upon facilitates the systematic improvement and iterative refinement of psychological theories, thus substantially increasing the efficiency of research.
While FAIR theory does not directly reduce ambiguity,
it provides a framework within which scholars can iteratively increase precision and formalization.
The FAIR principles also facilitate new ways of collaboration,
leveraging tools like Git for version control and Zenodo for archiving to document provenance and facilitate contributions from diverse researchers.
<!-- These practices align with the open science ethos, ensuring theories evolve dynamically based on empirical evidence and collaborative input. -->

>>>>

The so-called "replication crisis" has prompted extensive scientific reforms [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that the abundance of non-replicable findings 
was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses by commiting to an analysis plan in advance.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so ambiguous that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
The difference between these definitions becomes clear when considering the act of changing an infant's dirty diaper.
Under the original definition, one might hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
Under the expanded definition, one might hypothesize the opposite, as the act may fulfill the need for relatedness.

Scholars have raised concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
One main concern is that theories lack *formalization* [@szollosiArrestedTheoryDevelopment2021].
<!-- I still believe formalization != precision. Theories can be 100% formal and still 0% precise. -->
<!-- NvD: We should add somewhere what we mean by "formalization." In philosophy of science, formal logic about syntax (the propositions and predicates can have any meaning). However, what we mean, I think, is the explication of a narrative in (something so detailed and explicit that it can be captured with little error in) analytical or computational form (but the meaning remains. E, M, and C stand form something is E = MC<sup>2</sup>).   -->
When theories are ambiguous and hence require either subjective interpretation or additional background knowledge,
it becomes difficult do derive precise predictions,
and therefore hard to falsify the theory. 
A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
Despite these concerns, scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and improvement have been largely overlooked^[Here, we use "induction" to refer to inferences from specific observations to general theories. Others have used the term "abduction" to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960]. For present purposes, the terms are interchangeable.].
<!-- CvL considers abduction to be a special case of induction, because defining what makes a theory "best" or "plausible" requires making auxiliary assumptions, which restrict the space of potential inductive inferences. Peirce also stated that abduction "is not essentially different from induction" [@peirceCollectedPapersCharles1960, p.4881].  NvD dissents: -->
<!-- NvD: I cannot accept this footnote as an adequate characterization. Abduction is not just "inference to the best explanation." It is typically used as inference to a plausible explanation (but there could be more). See for instance Brian Haig's 2005 paper. Also, I don't see why it matters either way "why defining what makes a theory "best" requires auxiliary assumptions". At the very least I would like to change "we consider abduction to be a special case of induction", because I disagree. I am fine with the last sentence though ;-) My suggestion is to write the following: -->
<!-- NvD: We use induction to describe inferences from specific observations to general theories. In philosophy of science the term "abduction" is often used for the inference from observations to plausible explanations. We note that induction and abduction come in several varieties in philosophy science, which overlap in several respects For present purposes, however, the terms are interchangeable. -->
<!-- Peirce: CP 8.233 Cross-Ref:†† -->
<!-- 233. This is not essentially different from induction. Only it is most -->
<!-- usually an induction from instances which are not discrete and numerable. I now -->
<!-- call it Qualitative Induction. It is this which I used to confound with the second -->
<!-- line of procedure, or at least not to distinguish it sharply. -->
The present paper addresses this knowledge gap by applying, for the first time,
open science principles to psychological theory.
We apply introduce *FAIR theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use. 
<!-- NvD: Maybe we should add "specification" as part of "formalization" or "formalisation" as a part of "specification", because not everybody that is advocating better theories thinks that we should focus on formalization.  -->
More rigorous formalization increases theories' falsifiability [@popperLogicScientificDiscovery2002] because it expresses ideas as specific statements, clearly demarcating what should (not) be observed if the theory were true.
For example, Baddeley's verbal description of the phonological loop in his theory of working memory stands out for clarity and comprehensibility, yet it allows for at least 144 different implementations depending on the specification of various parameters such as decay rate, recall success, or rehearsal sequence, which were left undefined in the original theory [@lewandowsky2010computational].
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Compared to theories expressed in natural language,
formal theories facilitate inconsistency checking and evaluation of a theory's (lack of) vagueness.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.

@@

Making theory FAIR allows researchers to more easily find a relevant framework;
access and understand it; interact with it in a practical sense, for example, by deriving predictions from it, or using it to select control variables; and reuse it, contributing changes to existing theories or splitting of in a new direction.
Whereas the idea of theory can be quite nebulous to empirical psychologists,
FAIR theory makes theoretical work practical and tangible, incorporating theory into scholars' workflows.
Having a concrete object to iterate upon facilitates the systematic improvement and iterative refinement of psychological theories, thus substantially increasing the efficiency of research.
While FAIR theory does not directly reduce vagueness,
it provides a framework within which scholars can iteratively increase precision and formalization.
The FAIR principles also facilitate new ways of collaboration,
leveraging tools like Git for version control and Zenodo for archiving to document provenance and facilitate contributions from diverse researchers.
<!-- These practices align with the open science ethos, ensuring theories evolve dynamically based on empirical evidence and collaborative input. -->
